---
title: "MSADS 502 Final Project Code"
author: "Tommy"
date: "2024-12-07"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(car)
library(MLmetrics)
library(rpart)
library(rpart.plot)
library(C50)
library(zoo)
library(randomForest)
library(nnet)
library(NeuralNetTools)
library(e1071)
library(pROC)
library(ggcorrplot)
library(tidyr)
library(psych)
```

Upload data and check for N/A values

```{r}
df <- read.csv("us_stock_data.csv")
head(df)
colSums(is.na(df))
```

# Data Pre-processing

## Reformat date using lubridate package

```{r}
df$standardized_dates <- parse_date_time(df$Date, orders = c("dmy"))
df$standardized_dates <- as.Date(df$standardized_dates)
head(df)
```

We will be focusing on predicting Bitcoin Price Movement using individual stocks

Individual stocks: Apple, Tesla, Microsoft, Google, Nvidia, Berkshire Hathaway, Netflix, Amazon, and Meta

```{r}
# Create filtered data frame with individual stocks and dates

filtered_df <- df[, c("standardized_dates", "Bitcoin_Price", "Apple_Price", "Tesla_Price", "Microsoft_Price", "Google_Price", "Nvidia_Price", "Berkshire_Price", "Netflix_Price", "Amazon_Price", "Meta_Price")]

filtered_df$Bitcoin_Price <- gsub("[^0-9.-]", "", filtered_df$Bitcoin_Price)
filtered_df$Bitcoin_Price <- as.numeric(filtered_df$Bitcoin_Price)
filtered_df$Berkshire_Price <- gsub("[^0-9.-]", "", filtered_df$Berkshire_Price)
filtered_df$Berkshire_Price <- as.numeric(filtered_df$Berkshire_Price)
head(filtered_df)

# Sort the filtered data frame by date

sorted_filtered_df <- filtered_df[order(filtered_df$standardized_dates, decreasing = TRUE),]
head(sorted_filtered_df)
```

## Partition the Data into Test (25%) and Train (75%)

```{r}
split_point <- floor(0.25 * nrow(sorted_filtered_df))

test_data <- sorted_filtered_df[1:split_point,]
train_data <- sorted_filtered_df[(split_point + 1):nrow(sorted_filtered_df), ]

n_test <- dim(test_data)[1]
test_data$Index <- c(1:n_test)
n_train <- dim(train_data)[1]
train_data$Index <- c(1:n_train)

# Uncomment to check the test and train data

## Train data information:
## 760 Data points from 2020-01-02 to 2023-01-31

## Test data information:
## 253 Data points from 2023-02-01 to 2024-02-02

head(train_data)
head(test_data)
```

## Create Test & Train sets that have standardized variables

```{r}
# Standardize the variables into a new data frame

std_sorted_filtered_df <- sorted_filtered_df
std_sorted_filtered_df[, 2:11] <- scale(sorted_filtered_df[, 2:11])

# Split the new standardized data frame into the train and test sets

std_split_point <- floor(0.25 * nrow(std_sorted_filtered_df))

std_test <- std_sorted_filtered_df[1:std_split_point,]
std_train <- std_sorted_filtered_df[(split_point + 1):nrow(std_sorted_filtered_df),]

# Create Index for the train and test sets

n_test <- dim(std_test)[1]
std_test$Index <- c(1:n_test)
n_train <- dim(std_train)[1]
std_train$Index <- c(1:n_train)

# For time series analysis lag the target variable by one day to use previous day predictor variables to predict the movement of next day target variable

std_train$Bitcoin_Lag <- lag(std_train$Bitcoin_Price, 1)
std_test$Bitcoin_Lag <- lag(std_test$Bitcoin_Price, 1)

## Create a seven day rolling mean for the target variable

std_train <- std_train %>%
  mutate(var1_7day_avg = rollmean(Bitcoin_Price, k=7, fill = NA, align = "right"))
std_test <- std_test %>%
  mutate(var2_7day_avg = rollmean(Bitcoin_Price, k=7, fill = NA, align = "right"))

## Use the seven day rolling mean to see if the predictor variable increases or decreases

std_train <- std_train %>%
  mutate(Bitcoin_Price_Change = ifelse(var1_7day_avg > lag(var1_7day_avg), "Increased",
                                       ifelse(var1_7day_avg < lag(var1_7day_avg), "Decreased", "No Change")))
std_test <- std_test %>%
  mutate(Bitcoin_Price_Change = ifelse(var2_7day_avg > lag(var2_7day_avg), "Increased",
                                ifelse(var2_7day_avg < lag(var2_7day_avg), "Decreased", "No Change")))

std_train$Bitcoin_Price_Change <- factor(std_train$Bitcoin_Price_Change)
std_test$Bitcoin_Price_Change <- factor(std_test$Bitcoin_Price_Change)

std_train <- na.omit(std_train)
std_test <- na.omit(std_test)

head(std_train)
head(std_test)

table(std_train$Bitcoin_Price_Change)
table(std_test$Bitcoin_Price_Change)
```

# Exploratory Data Analysis

## Overlapping Line Graph with Stock Prices

Note: Berkshire_Price was not graphed because it has extreme values

```{r}
plot(train_data$standardized_dates, train_data$Apple_Price, type = "l", col = "#648fff", lwd = 3, xlab = "Dates over time", ylab = "Stock Prices", main = "Stock Prices Over Time From Jan 2020 to Jan 2023", ylim = c(0, 750))
lines(train_data$standardized_dates, train_data$Tesla_Price, col = "#785ef0", lwd = 3)
lines(train_data$standardized_dates, train_data$Microsoft_Price, col = "#dc267f", lwd = 3)
lines(train_data$standardized_dates, train_data$Google_Price, col = "#fe6100", lwd = 3)
lines(train_data$standardized_dates, train_data$Nvidia_Price, col = "#ffb000", lwd = 3)
# lines(train_data$standardized_dates, train_data$Berkshire_Price, col = "#785ef0", lwd = 2)
lines(train_data$standardized_dates, train_data$Netflix_Price, col = "#882255", lwd = 3)
lines(train_data$standardized_dates, train_data$Amazon_Price, col = "#0072b2", lwd = 3)
lines(train_data$standardized_dates, train_data$Meta_Price, col = "#117733", lwd = 3)
legend("topright", legend = c("Apple", "Tesla", "Microsoft", "Google", "Nvidia", "Netflix", "Amazon", "Meta"),col = c("#648fff", "#785ef0", "#dc267f", "#fe6100", "#ffb000", "#882255", "#0072b2", "#117733"), lwd = 10, cex = 0.65)
```

## Overlapping Line Graph with Standardized Stock Prices (Berkshire Included)

```{r}
plot(std_train$standardized_dates, std_train$Apple_Price, type = "l", col = "#648fff", lwd = 3, xlab = "Dates over time", ylab = "Stock Prices", main = "Stock Prices Over Time From Jan 2020 to Jan 2023", ylim = c(-3,3))
lines(std_train$standardized_dates, std_train$Tesla_Price, col = "#785ef0", lwd = 3)
lines(std_train$standardized_dates, std_train$Microsoft_Price, col = "#dc267f", lwd = 3)
lines(std_train$standardized_dates, std_train$Google_Price, col = "#fe6100", lwd = 3)
lines(std_train$standardized_dates, std_train$Nvidia_Price, col = "#ffb000", lwd = 3)
lines(std_train$standardized_dates, std_train$Berkshire_Price, col = "#000000", lwd = 2)
lines(std_train$standardized_dates, std_train$Netflix_Price, col = "#882255", lwd = 3)
lines(std_train$standardized_dates, std_train$Amazon_Price, col = "#0072b2", lwd = 3)
lines(std_train$standardized_dates, std_train$Meta_Price, col = "#117733", lwd = 3)
legend("topright", legend = c("Apple", "Tesla", "Microsoft", "Google", "Nvidia", "Berkshire", "Netflix", "Amazon", "Meta"),col = c("#648fff", "#785ef0", "#dc267f", "#fe6100", "#ffb000", "#000000", "#882255", "#0072b2", "#117733"), lwd = 10, cex = 0.65)
```

## Overlapping Line Graph with Only Apple, Bitcoin, Microsoft, and Tesla

```{r}
# Convert wide data to long format for visualization
std_train_long <- std_train %>%
  pivot_longer(
    cols = c(Bitcoin_Price, Apple_Price, Tesla_Price, Microsoft_Price, 
    ), 
    names_to = "Stock", 
    values_to = "Price"
  )

# Create line plots for all stocks and Bitcoin price over time
ggplot(std_train_long, aes(x = standardized_dates, y = Price, color = Stock)) +
  geom_line() +
  labs(title = "Stock and Bitcoin Price Trends Over Time",
       x = "Date",
       y = "Price",
       color = "Stock") +
  theme_minimal()
```

## Overlapping Line Graph with Only Berkshire, Bitcoin, Google, and Nvidia

```{r}
long_data <- std_train %>%
  pivot_longer(cols = c(Bitcoin_Price, Google_Price, Nvidia_Price, Berkshire_Price), 
               names_to = "Stock", values_to = "Price")

ggplot(long_data, aes(x = standardized_dates, y = Price, color = Stock)) +
  geom_line() +
  labs(title = "Price Trends Over Time", x = "Date", y = "Price") +
  theme_minimal()
```

## Overlapping Line Graph with Only Amazon, Bitcoin, Meta, and Netflix

```{r}
long_data <- std_train %>%
  pivot_longer(cols = c(Bitcoin_Price, Netflix_Price, 
                        Amazon_Price, Meta_Price), 
               names_to = "Stock", values_to = "Price")

ggplot(long_data, aes(x = standardized_dates, y = Price, color = Stock)) +
  geom_line() +
  labs(title = "Price Trends Over Time", x = "Date", y = "Price") +
  theme_minimal()
```

## Correlation Heat Map of all Variables

```{r}
#Correlation Analysis
#Explore the correlation between Bitcoin's price and the prices of individual stocks

correlation_matrix <- cor(train_data[, 2:11], use = "complete.obs")
print(correlation_matrix)

# Heatmap of correlations
ggcorrplot(correlation_matrix, lab = TRUE)
```

```{r}
#Seasonality or Temporal Trends
#Check if there are any temporal patterns in Bitcoin price changes

ggplot(std_train, aes(x = month(standardized_dates), fill = Bitcoin_Price_Change)) +
  geom_bar() +
  labs(title = "Bitcoin Price Change by Month", x = "Month", y = "Frequency") +
  theme_minimal()
```

# Model building

## Logistic Regression Model Based on Stock Prices

Accuracy: (39 + 81)/246 = 0.4878
Error Rate: 1 - 0.4878 = 0.5122
Sensitivity: 81/113 = 0.7168
Specificity: 39/133 = 0.3451
Precision: 81/175 = 0.4629
F1: (2)(0.4629)(0.7168)/(0.4629 + 0.7168) = 0.5625
F2: (5)(0.4629)(0.7168)/((4)(0.4629) + 0.7168) = 0.6459
F0.5: (1.25)(0.4629)(0.7168)/((0.25)(0.4629) + 0.7168) = 0.4982

```{r}
lrs01_model <- glm(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, family = binomial)
summary(lrs01_model)
print("Predictor Variable Multicollinearity:")

ypred <- predict(object = lrs01_model, newdata = std_test, type = "response")
predicted_classes <- ifelse(ypred > 0.7, 1, 0)

# Contingency Table

t1 <- table(std_test$Bitcoin_Price_Change, predicted_classes)
row.names(t1) <- c("Actual: Decrease", "Actual: Increase")
colnames(t1) <- c("Predicted: Decrease", "Predicted: Increase")
t1 <- addmargins(A = t1, FUN = list(Total = sum), quiet = TRUE)
t1

# Logistic Regression AUC score

lrs01_actual <- (std_test$Bitcoin_Price_Change == "Increased")
lrs01_roc <- roc(lrs01_actual, predicted_classes)
lrs01_auc <- auc(lrs01_roc)
lrs01_auc
```

## Logistic Regression Model with Removed Predictor Variables

Accuracy: (23 + 100)/246 = 0.5000
Error Rate: 1 - 0.5000 = 0.5000
Sensitivity: 100/113 = 0.8850
Specificity: 23/133 = 0.1729
Precision: 100/210 = 0.4762
F1: (2)(0.4762)(0.8850)/(0.4762 + 0.8850) = 0.6192
F2: (5)(0.4762)(0.8850)/((4)(0.4762) + 0.8850) = 0.7553
F0.5: (1.25)(0.4762)(0.8850)/((0.25)(0.4762) + 0.8850) = 0.5247

```{r}
lrs02_model <- glm(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Berkshire_Price, data = std_train, family = binomial)
summary(lrs02_model)
print("Predictor Variable Multicollinearity:")
vif(lrs02_model)

ypred <- predict(object = lrs02_model, newdata = std_test, type = "response")
predicted_classes <- ifelse(ypred > 0.5, 1, 0)

# Contingency Table

t2 <- table(std_test$Bitcoin_Price_Change, predicted_classes)
row.names(t2) <- c("Actual: Decrease", "Actual: Increase")
colnames(t2) <- c("Predicted: Decrease", "Predicted: Increase")
t2 <- addmargins(A = t2, FUN = list(Total = sum), quiet = TRUE)
t2

# Logistic Regression AUC score

lrs02_actual <- (std_test$Bitcoin_Price_Change == "Increased")
lrs02_roc <- roc(lrs02_actual, predicted_classes)
lrs02_auc <- auc(lrs02_roc)
lrs02_auc
```

## CART

Accuracy: (83 + 55)/246 = 0.5610
Error Rate: 1 - 0.5610 = 0.4390
Sensitivity: 55/113 = 0.4867
Specificity: 83/133 = 0.6241
Precision: 55/105 = 0.5238
F1: (2)(0.5238)(0.4867)/(0.5238 + 0.4867) = 0.5046
F2: (5)(0.5238)(0.4867)/((4)(0.5238) + 0.4867) = 0.4937
F0.5: (1.25)(0.5238)(0.4867)/((0.25)(0.5238) + 0.4867) = 0.5159

```{r}
cart01_model <- rpart(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, method = "class")

rpart.plot(cart01_model)

ypred <- predict(object = cart01_model, newdata = std_test, type = "class")
# predicted_class <- ifelse(ypred[, 2] > 0.5, 1, 0)

# Contingency Table

t1 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t1) <- c("Actual: Decrease", "Actual: Increase")
colnames(t1) <- c("Predicted: Decrease", "Predicted: Increase")
t1 <- addmargins(A = t1, FUN = list(Total = sum), quiet = TRUE)
t1

# Node counts

total_nodes <- length(cart01_model$frame$var)
total_nodes
leaf_nodes <- sum(cart01_model$frame$var == "<leaf>")
leaf_nodes
decision_nodes <- total_nodes - leaf_nodes
decision_nodes

# Information regarding the CART model CP score

# summary(cart01_model)$splits
```

## Pruned CART Model

Accuracy: (74 + 64)/246 = 0.5610
Error Rate: 1 - 0.5610 = 0.4390
Sensitivity: 64/113 = 0.5664
Specificity: 74/133 = 0.5564
Precision: 64/123 = 0.5203
F1: (2)(0.5203)(0.5664)/(0.5203 + 0.5664) = 0.5424
F2: (5)(0.5203)(0.5664)/((4)(0.5203) + 0.5664) = 0.5565
F0.5: (1.25)(0.5203)(0.5664)/((0.25)(0.5203) + 0.5664) = 0.5289

```{r}
# cart01_model$cptable
cart02_model <- prune(cart01_model, cp = 0.015)

rpart.plot(cart02_model)

ypred <- predict(object = cart02_model, newdata = std_test, type = "prob")
predicted_classes <- ifelse(ypred[, 2] > 0.3, 1, 0)

# Contingency Table

t2 <- table(std_test$Bitcoin_Price_Change, predicted_classes)
row.names(t2) <- c("Actual: Decrease", "Actual: Increase")
colnames(t2) <- c("Predicted: Decrease", "Predicted: Increase")
t2 <- addmargins(A = t2, FUN = list(Total = sum), quiet = TRUE)
t2

# Node counts

total_nodes <- length(cart02_model$frame$var)
total_nodes
leaf_nodes <- sum(cart02_model$frame$var == "<leaf>")
leaf_nodes
decision_nodes <- total_nodes - leaf_nodes
decision_nodes

# CART AUC score

cart02_actual <- (std_test$Bitcoin_Price_Change == "Increased")
cart02_roc <- roc(cart02_actual, predicted_classes)
cart02_auc <- auc(cart02_roc)
cart02_auc
```

## Random Forests

Accuracy: (41 + 92)/246 = 0.5407
Error Rate: 1 - 0.4919 = 0.4593
Sensitivity: 92/113 = 0.8142
Specificity: 41/133 = 0.3083
Precision: 92/184 = 0.4457
F1: (2)(0.4457)(0.8142)/(0.4457 + 0.8142) = 0.5761
F2: (5)(0.4457)(0.8142)/((4)(0.4457) + 0.8142) = 0.6987
F0.5: (1.25)(0.4457)(0.8142)/((0.25)(0.4457) + 0.8142) = 0.4901

```{r}
set.seed(35)
rf01_model <- randomForest(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, method = "class")

ypred <- predict(object = rf01_model, newdata = std_test, type = "class")

# Contingency Table

t1 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t1) <- c("Actual: Decrease", "Actual: Increase")
colnames(t1) <- c("Predicted: Decrease", "Predicted: Increase")
t1 <- addmargins(A = t1, FUN = list(Total = sum), quiet = TRUE)
t1

rf01_model$ntree

# RF AUC score

rf01_roc <- roc(std_test$Bitcoin_Price_Change, as.numeric(ypred))
rf01_auc <- auc(rf01_roc)
rf01_auc
```

## Random Forests with 1000 ntree

```{r}
set.seed(35)
rf02_model <- randomForest(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, method = "class", ntree = 1000)

ypred <- predict(object = rf02_model, newdata = std_test, type = "class")

# Contingency Table

t2 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t2) <- c("Actual: Decrease", "Actual: Increase")
colnames(t2) <- c("Predicted: Decrease", "Predicted: Increase")
t2 <- addmargins(A = t2, FUN = list(Total = sum), quiet = TRUE)
t2

rf02_model$ntree

# RF AUC score

rf02_roc <- roc(std_test$Bitcoin_Price_Change, as.numeric(ypred))
rf02_auc <- auc(rf02_roc)
rf02_auc
```

## Random Forests with 100 ntree

```{r}
set.seed(35)
rf03_model <- randomForest(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, method = "class", ntree = 100)

ypred <- predict(object = rf03_model, newdata = std_test, type = "class")

# Contingency Table

t3 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t3) <- c("Actual: Decrease", "Actual: Increase")
colnames(t3) <- c("Predicted: Decrease", "Predicted: Increase")
t3 <- addmargins(A = t3, FUN = list(Total = sum), quiet = TRUE)
t3

rf03_model$ntree

# RF AUC score

rf03_roc <- roc(std_test$Bitcoin_Price_Change, as.numeric(ypred))
rf03_auc <- auc(rf03_roc)
rf03_auc
```

## C5.0

```{r}
c5_model <- C5.0(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train, methond = "class")

plot(c5_model)

ypred <- predict(object = c5_model, newdata = std_test, type = "class")

t1 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t1) <- c("Actual: Decrease", "Actual: Increase")
colnames(t1) <- c("Predicted: Decrease", "Predicted: Increase")
t1 <- addmargins(A = t1, FUN = list(Total = sum), quiet = TRUE)
t1
```

## Neural Networks

```{r}
set.seed(35)
nnet01_model <- nnet(formula = Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + 
                     Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + 
                     Netflix_Price + Amazon_Price + Meta_Price, data = std_train, size = 1)

plotnet(nnet01_model)

ypred <- predict(object = nnet01_model, newdata = std_test, type = "class")

t1 <- table(std_test$Bitcoin_Price_Change, ypred)
row.names(t1) <- c("Actual: Decrease", "Actual: Increase")
colnames(t1) <- c("Predicted: Decrease", "Predicted: Increase")
t1 <- addmargins(A = t1, FUN = list(Total = sum), quiet = TRUE)
t1
```

## Naive Bayes

```{r}
# Train the Naive Bayes model
nb_model <- naiveBayes(Bitcoin_Price_Change ~ Apple_Price + Tesla_Price + Microsoft_Price + Google_Price + Nvidia_Price + Berkshire_Price + Netflix_Price + Amazon_Price + Meta_Price, data = std_train)

# Predict on the test set
nb_predictions_prob <- predict(nb_model, newdata = std_test, type = "raw")
nb_predictions <- ifelse(nb_predictions_prob[, "Increased"] >= 0.9, "Increased", "Decreased")

# Confusion matrix
conf_matrix <- table(std_test$Bitcoin_Price_Change, nb_predictions)

# Add proper row and column names
row.names(conf_matrix) <- c("Actual: Decrease", "Actual: Increase")
colnames(conf_matrix) <- c("Predicted: Decrease", "Predicted: Increase")

# Print the confusion matrix
print(conf_matrix)
```